Optimal forward modeling of gravitational fields using tesseroids with adaptive discretization

### 2014-08-20 manuscript.tex
Moved manuscript to GJI template

%\documentclass[manuscript]{geophysics}
%\usepackage[nomarkers]{endfloat}
% For a two-column format that looks like the final product, uncomment bellow
% and comment the two lines above
\documentclass[paper,twocolumn,twoside]{geophysics}

\usepackage{graphicx}
\usepackage{amsmath}

\newcommand{\vect}[1]{\mathbf{#1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\comp}[1]{#1^{\alpha\beta}}
\newcommand{\norm}[1]{\left|\left|#1\right|\right|}

@@

\title{
    Tesseroids:
    command-line tools for
    gravity field forward modeling
    in spherical coordinates
}

% manuscript number
\ms{}

\address{
    \footnotemark[1]
    Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil,
    e-mail: leouieda@gmail.com;\\
    \footnotemark[2]
    Observat\'orio Nacional, Rio de Janeiro, Brazil;
}

\author{Leonardo Uieda\footnotemark[1]\footnotemark[2],
        Vanderlei C. Oliveira Jr\footnotemark[2],
        and
        Val\'eria C. F. Barbosa\footnotemark[2]}

\lefthead{Uieda et al.}
\righthead{Gravity forward modeling with Tesseroids}

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nam eu dolor pretium,
egestas mauris sed, dapibus quam. Duis hendrerit mollis nunc a consequat. Nulla
et sem consectetur, interdum velit eget, aliquam ipsum. Praesent sagittis
tortor diam, sed ultrices magna ullamcorper vitae. Proin vitae orci augue.
Morbi dictum ligula gravida sem malesuada facilisis. Mauris nibh metus, cursus
eget imperdiet vitae, pretium at lorem. Praesent nisi mauris, pretium ut risus
fermentum, egestas tincidunt nibh. Mauris nulla orci, consequat eu pharetra
non, mattis ut urna. Mauris facilisis orci eros. Nam mattis non magna iaculis
consectetur. Morbi sodales dolor vitae felis sagittis, eget faucibus turpis
convallis. Nullam malesuada, mauris et ultricies rutrum, odio nulla gravida
nunc, ac volutpat eros lectus eget lacus. Integer venenatis velit vel justo
pellentesque, quis molestie sem vestibulum.
\end{abstract}

@@

\noindent
in which
$(\phi, \lambda, r)$ are
the latitude, longitude, and radius
coordinates of computation point P (Figure~\ref{fig:tesseroid}),
$\alpha,\beta \in \{x,y,z\}$
(the local coordinate system of point P),
and

@@

We use a modified version
of the recursive scheme of \citet{Li2011}.
We fix the number of nodes in the GLQ
$N^\lambda=N^\phi=N^r=2$.
In this case, the maximum distance between nodes
is approximately the largest dimension of the tesseroid ($L$)
\citep{Wild-Pfeiffer2008}.
For a computation point $P$
at a distance $d$ from
the center of the top face of the tesseroid
(Figure~\ref{fig:ratio}a),
the tesseroid is divided in eight parts if

@@

The rule-of-thumb of \citet{Ku1977}
used by \citet{Li2011}
is equivalent to setting $R=1$.
Figure~\ref{fig:ratio}
illustrates the subdivision of a tesseroid
(Figure~\ref{fig:ratio}a)
for computation point $P$
using increasing values of $R$.
Note that increasing $R$
results is a finer division of the tesseroid
around the computation point
and thus a more accurate GLQ integration.
However, higher values of $R$
also result in
larger computation times.
For example,
the tesseroid in Figure~\ref{fig:ratio}a
is replaced by
50 tesseroids for $R=0.5$ (Figure~\ref{fig:ratio}b),
344 tesseroids for $R=1$ (Figure~\ref{fig:ratio}c),
and 2472 tesseroids for  $R=2$ (Figure~\ref{fig:ratio}d).

@@

Clearly,
there is need to determine
an optimal value of $R$
that yields the desired accuracy
in the GLQ integration.
In the following section,
we compare the computed gravitational effects
of a tesseroid model
with that of a spherical half-shell
(Figure~\ref{fig:shell}),
for which exist analytical expressions.

@@

The gravitational potential,
gravitational attraction,
and
Marussi (gravity gradient) tensor
caused by the homogeneous spherical half-shell
shown in Figure~\ref{fig:shell}
at a computation point
$P = (\lambda=90^\circ, \phi=0^\circ, r=r_2+h)$
are

\begin{equation}
    V(h) = 2\pi G \rho \left[
        \dfrac{l^3 + {r'}^3}{3(r_2 + h)} - 0.5 {r'}^2 \right]
         \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellpot}
\end{equation}
\begin{equation}
    g_z(h) = 2\pi G \rho \left[
        \dfrac{l^3 + {r'}^3}{3(r_2 + h)^2} - l \right]
        \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellgz}
\end{equation}
\begin{equation}
    g_{xx}(h) = g_{yy}(h) = -\dfrac{1}{2} g_{zz}(h),
\end{equation}
\begin{equation}
    g_{xy}(h) = g_{xz}(h) = g_{yz}(h) = 0,
\end{equation}

@@

\begin{equation}
    g_{zz}(h) = 2\pi G \rho \left[
        2\dfrac{l^3 + {r'}^3}{3(r_2 + h)^3}
        - \dfrac{l}{r_2 + h} + \dfrac{r_2 + h}{l}
        \right]
        \Biggr \rvert_{r'=r_1}^{r'=r_2},
    \label{eq:shellgzz}
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/error-20deg}
    \caption{This is a figure caption.}
    \label{fig:error20}
\end{figure}

This is a reference to Figure \ref{fig:error20}.

@@

\bibliographystyle{seg}
\bibliography{references}

>>>>

%\documentclass[extra,mreferee]{gji}
% Use the option below to make a nicely formatted manuscript
\documentclass[extra]{gji}
\usepackage{timet}
\usepackage{graphicx}

\title[Optimal forward modeling using tesseroids]{
    Optimal forward modeling of gravitational fields
    using tesseroids with adaptive discretization
}
\author[Uieda et al.]{
    Leonardo Uieda$^{1,2}$,
    Vanderlei C. Oliveira Jr$^2$,
    Val\'eria C. F. Barbosa$^2$
    \\
    $^1$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil.
    $^2$Observat\'orio Nacional, Rio de Janeiro, Brazil.
}

@@

\label{firstpage}
\maketitle

\begin{summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Nam eu dolor pretium,
egestas mauris sed, dapibus quam. Duis hendrerit mollis nunc a consequat. Nulla
et sem consectetur, interdum velit eget, aliquam ipsum. Praesent sagittis
tortor diam, sed ultrices magna ullamcorper vitae. Proin vitae orci augue.
Morbi dictum ligula gravida sem malesuada facilisis. Mauris nibh metus, cursus
eget imperdiet vitae, pretium at lorem. Praesent nisi mauris, pretium ut risus
fermentum, egestas tincidunt nibh. Mauris nulla orci, consequat eu pharetra
non, mattis ut urna. Mauris facilisis orci eros. Nam mattis non magna iaculis
consectetur. Morbi sodales dolor vitae felis sagittis, eget faucibus turpis
convallis. Nullam malesuada, mauris et ultricies rutrum, odio nulla gravida
nunc, ac volutpat eros lectus eget lacus. Integer venenatis velit vel justo
pellentesque, quis molestie sem vestibulum.
\end{summary}

@@

\noindent
in which
$(\phi, \lambda, r)$ are
the latitude, longitude, and radius
coordinates of computation point P (Fig.~\ref{fig:tesseroid}),
$\alpha,\beta \in \{x,y,z\}$
(the local coordinate system of point P),
and

@@

We use a modified version
of the recursive scheme of \citet{Li2011}.
We fix the number of nodes in the GLQ
$N^\lambda=N^\phi=N^r=2$.
In this case, the maximum distance between nodes
is approximately the largest dimension of the tesseroid ($L$)
\citep{Wild-Pfeiffer2008}.
For a computation point $P$
at a distance $d$ from
the center of the top face of the tesseroid
(Fig.~\ref{fig:ratio}a),
the tesseroid is divided in eight parts if

@@

The rule-of-thumb of \citet{Ku1977}
used by \citet{Li2011}
is equivalent to setting $R=1$.
Fig.~\ref{fig:ratio}
illustrates the subdivision of a tesseroid
(Fig.~\ref{fig:ratio}a)
for computation point $P$
using increasing values of $R$.
Note that increasing $R$
results is a finer division of the tesseroid
around the computation point
and thus a more accurate GLQ integration.
However, higher values of $R$
also result in
larger computation times.
For example,
the tesseroid in Fig.~\ref{fig:ratio}a
is replaced by
50 tesseroids for $R=0.5$ (Fig.~\ref{fig:ratio}b),
344 tesseroids for $R=1$ (Fig.~\ref{fig:ratio}c),
and 2472 tesseroids for  $R=2$ (Fig.~\ref{fig:ratio}d).

@@

Clearly,
there is need to determine
an optimal value of $R$
that yields the desired accuracy
in the GLQ integration.
In the following section,
we compare the computed gravitational effects
of a tesseroid model
with that of a spherical half-shell
(Fig.~\ref{fig:shell}),
for which exist analytical expressions.

@@

The gravitational potential,
gravitational attraction,
and
Marussi (gravity gradient) tensor
caused by the homogeneous spherical half-shell
shown in Fig.~\ref{fig:shell}
at a computation point
$P = (\lambda=90^\circ, \phi=0^\circ, r=r_2+h)$
are

\begin{equation}
    V(h) = 2\pi G \rho \left[
        \frac{l^3 + {r'}^3}{3(r_2 + h)} - 0.5 {r'}^2 \right]
         \Biggr|_{r'=r_1}^{r'=r_2},
    \label{eq:shellpot}
\end{equation}
\begin{equation}
    g_z(h) = 2\pi G \rho \left[
        \frac{l^3 + {r'}^3}{3(r_2 + h)^2} - l \right]
        \Biggr|_{r'=r_1}^{r'=r_2},
    \label{eq:shellgz}
\end{equation}
\begin{equation}
    g_{xx}(h) = g_{yy}(h) = -\frac{1}{2} g_{zz}(h),
\end{equation}
\begin{equation}
    g_{xy}(h) = g_{xz}(h) = g_{yz}(h) = 0,
\end{equation}

@@

\begin{equation}
    g_{zz}(h) = 2\pi G \rho \left[
        2\frac{l^3 + {r'}^3}{3(r_2 + h)^3}
        - \frac{l}{r_2 + h} + \frac{r_2 + h}{l}
        \right]
        \Biggr|_{r'=r_1}^{r'=r_2},
    \label{eq:shellgzz}
\end{equation}

@@

Meh.

@@

\bibliographystyle{gji}
\bibliography{references}
